---
title: "7 - Moving Beyond Linearity"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r, message=FALSE}
# load libraries
library(tidymodels)
library(ISLR)

# load data
Wage <- as_tibble(Wage)
```

## 7.1 - Polynomial Regression and Step Functions {.tabset}

```{r}
# specify recipe
rec_poly <- recipe(wage ~ age, data = Wage) %>%
  step_poly(age, degree = 4)

# specify model
lm_spec <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")

# create workflow object
poly_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_recipe(rec_poly)

# fit model
poly_fit <- poly_wf %>% fit(Wage)

# get coefficients
tidy(poly_fit)
```

**Note**: `step_poly()` returns variables that are a basis of orthogonal polynomials, which means that each of the columns is a linear combination of the variables `age`, `age^2`, `age^3`, and `age^4`.

### Orthogonal polynomials

```{r}
# orthogonal polynomials (default)
poly(1:6, degree = 4, raw = FALSE)
```

### Raw polynomials

```{r}
# raw polynomials
poly(1:6, degree = 4, raw = TRUE)
```

## {.unlisted .unnumbered .tabset}

It is still recommended to stick with the default of `raw = FALSE` unless you have a reason not to do that. One of the benefits of using `raw = FALSE` is that the resulting variables are uncorrelated which is a desirable quality when using a linear regression model.

```{r}
# specify recipe (using raw polynomials)
rec_raw_poly <- recipe(wage ~ age, data = Wage) %>%
  step_poly(age, degree = 4, options = list(raw = TRUE))

# create workflow object
raw_poly_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_recipe(rec_raw_poly)

# fit model
raw_poly_fit <- raw_poly_wf %>% fit(Wage)

# output coefficients
tidy(raw_poly_fit)
```

```{r}
# get range of ages in the data set
age_range <- tibble(age = seq(min(Wage$age), max(Wage$age)))

# compute predictions and confidence bounds for the given ages
regression_lines <- bind_cols(
  augment(poly_fit, new_data = age_range),
  predict(poly_fit, new_data = age_range, type = "conf_int")
)

regression_lines
```

```{r}
# visualize the fitted line and confidence interval
Wage %>%
  ggplot(aes(age, wage)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = .pred),
            color = "darkgreen",
            data = regression_lines) +
  geom_line(aes(y = .pred_lower),
            color = "blue",
            linetype = "dashed",
            data = regression_lines) +
  geom_line(aes(y = .pred_upper),
            color = "blue",
            linetype = "dashed",
            data = regression_lines)
```

```{r}
# specify wider range of ages
wide_age_range <- tibble(age = seq(18, 100))

# compute predictions and confidence bounds for the given ages
regression_lines <- bind_cols(
  augment(poly_fit, new_data = wide_age_range),
  predict(poly_fit, new_data = wide_age_range, type = "conf_int")
)

# visualize the fitted line and confidence interval
Wage %>%
  ggplot(aes(age, wage)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = .pred),
            color = "darkgreen",
            data = regression_lines) +
  geom_line(aes(y = .pred_lower),
            color = "blue",
            linetype = "dashed",
            data = regression_lines) +
  geom_line(aes(y = .pred_upper),
            color = "blue",
            linetype = "dashed",
            data = regression_lines)
```

Classification example

```{r}
# change response to be a binary variable
Wage <- Wage %>%
  mutate(high = factor(wage > 250,
                       levels = c(TRUE, FALSE),
                       labels = c("High", "Low")))

# specify recipe
rec_poly <- recipe(high ~ age, data = Wage) %>%
  step_poly(age, degree = 4)

# specify model
lr_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# create workflow object
lr_poly_wf <- workflow() %>%
  add_model(lr_spec) %>%
  add_recipe(rec_poly)

# fit model
lr_poly_fit <- lr_poly_wf %>% fit(Wage)

# compute predictions and confidence bounds for the given ages
regression_lines <- bind_cols(
  augment(lr_poly_fit, new_data = age_range, type = "prob"),
  predict(lr_poly_fit, new_data = age_range, type = "conf_int")
)

# visualize the fitted line and confidence interval
regression_lines %>%
  ggplot(aes(age)) +
  ylim(c(0, 0.2)) +
  geom_line(aes(y = .pred_High), 
            color = "darkgreen") +
  geom_line(aes(y = .pred_lower_High),
            color = "blue", 
            linetype = "dashed") +
  geom_line(aes(y = .pred_upper_High), 
            color = "blue", 
            linetype = "dashed") +
  geom_jitter(aes(y = (high == "High") / 5),
              data = Wage,
              shape = "|",
              height = 0,
              width = 0.2)
```

Step function

### `step_discretize()`

Convert a numeric variable into a factor variable with `n` bins, with each bin having the same number of points

```{r}
# specify recipe
rec_discretize <- recipe(high ~ age, data = Wage) %>%
  step_discretize(age, num_breaks = 4)

# create workflow object
discretize_wf <- workflow() %>%
  add_model(lr_spec) %>%
  add_recipe(rec_discretize)

# fit model
discretize_fit <- discretize_wf %>% fit(Wage)
discretize_fit
```

### `step_cut()`

Manually supply the breaks

```{r}
# specify recipe
rec_cut <- recipe(high ~ age, data = Wage) %>%
  step_cut(age, breaks = c(30, 50, 70))

# create workflow object
cut_wf <- workflow() %>%
  add_model(lr_spec) %>%
  add_recipe(rec_cut)

# fit model
cut_fit <- cut_wf %>% fit(Wage)
cut_fit
```

## 7.2 - Splines

```{r}
# specify recipe
rec_spline <- recipe(wage ~ age, data = Wage) %>%
  step_bs(age, options = list(knots = 25, 40, 60))

# create workflow object
spline_wf <- workflow() %>%
  add_model(lm_spec) %>%
  add_recipe(rec_spline)

# fit model
spline_fit <- fit(spline_wf, data = Wage)

# compute predictions and confidence bounds for the given ages
regression_lines <- bind_cols(
  augment(spline_fit, new_data = age_range),
  predict(spline_fit, new_data = age_range, type = "conf_int")
)

# visualize the fitted line and confidence interval
Wage %>%
  ggplot(aes(age, wage)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = .pred),
            color = "darkgreen",
            data = regression_lines) +
  geom_line(aes(y = .pred_lower),
            color = "blue",
            linetype = "dashed",
            data = regression_lines) +
  geom_line(aes(y = .pred_upper),
            color = "blue",
            linetype = "dashed",
            data = regression_lines)
```

## Credit

[ISLR tidymodel labs - Chapter 7](https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/07-moving-beyond-linearity.html)
